---
title: Assessing Health Facility Capacity for Neglected Tropical Diseases in a Refugee
  Settlement in Uganda.
author: Elvis Tamale^1,2,6^, Paddy Derrick Malinga^2^ , Mable Ssebaduka Nabweteme^2^,Celine
  Ahurira^2^,Conrad Makai^2^,   Kenneth Ssembuze^3^, Patricia Ajeru^4^, Patience Atuhaire^5^
date: "2025-12-20"
output:
  html_document:
    sansfont: Times New Roman
    monofont: Times New Roman
    fontsize: 11pt
    linestretch: 1.5
    theme: flatly
    highlight: tango
    mainfont: Times New Roman
    toc: true
    toc_float: true
    number_sections: true
    code_download: true
    df_print: paged
    self_contained: true
    smooth_scroll: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tinytex)
```

**Affiliations**

^1^St Mary’s Hospital Lacor, Gulu city, Uganda. 

^2^Mbarara University of Science and Technology, Mbarara, Uganda. 

^3^Medical Teams International, Portland, USA. 

^4^Kampala International University, Kampala, Uganda.

^5^Makerere University Johns Hopkins University Research Collaboration, Kampala, Uganda 

^6^University of Cambridge, Cambridge,UK.

**Corresponding author**

Tamale Elvis, St. Mary's Hospital, Lacor, Gulu city, Uganda.

**Corresponding author email**

elvistamale325@gmail.com. 
 
**Keywords**

Neglected Tropical Diseases, Refugee Settlement, Uganda, Primary Healthcare, Health Workers, Knowledge, Capacity, Schistosomiasis, Soil-Transmitted Helminths. 


## Abstract 
### Background  
Neglected Tropical Diseases (NTDs) pose a substantial global public health challenge, particularly in vulnerable refugee populations. In Uganda, millions are at risk, yet primary healthcare (PHC) facility capacity within refugee settlements to effectively manage these diseases remains underexplored. This study assessed infrastructure, diagnostic capabilities, and resource availability for NTD management in health facilities across Nakivale Refugee Settlement, Uganda. 

### Methods 
A quantitative assessment was conducted across 8 health facilities (6 Health Centre III, 2 Health Centre IV) in Nakivale Refugee Settlement. Data collection included facility type, bed capacity, dedicated laboratory services and personnel, availability of diagnostic tests and guiding tools, status of reporting systems, and essential medicines/laboratory supplies for soil-transmitted helminths (STH) and schistosomiasis. Descriptive statistics summarized findings. 

### Results  
The 8 assessed facilities (6 HCIII, 2 HCIV) averaged 23 and 50 beds, respectively. All facilities reported dedicated laboratories with technicians and microscopes. Diagnostic tests (microscopes, dyes, assay strips) and guiding tools (clinical guidelines, laboratory protocols, bench aids) were consistently available, though specific test types varied. Five out of eight (62.5%) facilities had Health Management Information System reporting forms. Essential medicines, Albendazole and Mebendazole, were available in most facilities (7 had both, 1 had Albendazole only). Common laboratory supplies (e.g., centrifuge tubes, syringes) were universal, but critical items like cellophane and Kato Katz templates were absent in all facilities. Most facilities were over 10 km from the town council. 

### Conclusion 
Health facilities in Nakivale Refugee Settlement possess foundational infrastructure for NTD management, including laboratories, microscopes, and essential medicines. However, significant gaps exist in specific diagnostic supplies (e.g., Kato Katz templates, cellophane) and reporting system completeness. Their predominantly remote location also highlights potential access challenges. 
Addressing these gaps through targeted provision of laboratory consumables, strengthening reporting, and considering geographical accessibility is crucial for enhancing PHC capacity to manage NTDs in this refugee setting. 
 
## Introduction 
Neglected tropical diseases (NTDs) are a group of 13 infections caused by parasitic worms, protozoa, or bacteria currently affecting over 1 billion people, with another 1 billion at risk of infection [1-15]. They strike the world’s poorest people with adverse effects on the health, well-being, and socioeconomic facets of one’s life [2, 16-19]. Africa bears a substantial portion of this global burden, accounting for over 600 million cases, particularly affecting poor rural communities [1-2, 20-21]. Uganda is endemic for all five NTDs, with more than 40 million people at risk for one or more NTDs targeted by the United States Agency for International Development (USAID) Act to End NTDs [22-23]. 
Despite their significant burden, NTDs are largely preventable and amenable to elimination. In 2020, the 
World Health Organization (WHO) outlined a roadmap for eradicating and eliminating 20 NTDs by 2030 [24-25]. The WHO, alongside policymakers and academics, advocates for integrating NTD management into existing public health programs within endemic countries [26-27]. Such integration is posited to contribute significantly to achieving several sustainable development goals (SDGs) and vice versa [24, 28-31]. Despite the burden of NTDs in Sub-Saharan Africa, NTDs have not been well integrated into the PHC of most countries. Previous studies in Nigeria, Tanzania, and Burundi have consistently reported low capacity and readiness among health workers to diagnose and manage NTDs [1, 32-35]. The Ministry of Health in Uganda has identified several hindrances to eliminating NTDs by 2030, including a lack of knowledge, skills, and equipment for diagnosis and management in health facilities and deficiencies in reporting systems [36]. 
Nakivale Refugee Settlement, the eighth-largest settlement globally, hosts approximately 145,613 refugees from ten different nationalities [37-38]. Within this settlement, NTDs, particularly schistosomiasis and soil-transmitted helminths (STHs), are prevalent, with reported rates of 26.6% and 26.5%, respectively [36]. The increasing number of refugees, coupled with inadequate resources, raises critical questions about the capacity of the existing health system to effectively diagnose, manage, and treat these NTDs. Moreover, comprehensive data on NTDs and their prevalence are scarce in regions such as southwestern Uganda. Uganda's national strategy for NTDs involves integrated control programmes [39]. This study assessed the capacity of primary healthcare centers to timely diagnose, treat, and manage soil-transmitted helminths and schistosomiasis among refugees in Nakivale Refugee Settlement, Southwestern Uganda, by looking at the availability of essential diagnostic and reporting infrastructure. 
 
 
## Methods 
### Study Design and Setting 
This study employed a quantitative, cross-sectional design focusing on assessing health facility capacity. The study was conducted in Nakivale Refugee Settlement, located in southwestern Uganda, a large settlement hosting a diverse refugee population. 

### Study Population and Sampling 
The study population comprised 8 health facilities within Nakivale Refugee Settlement. These facilities included 6 Health Centre III (HCIII) and 2 Health Centre IV (HCIV) facilities, representing key points of primary healthcare service delivery in the settlement. These facilities were purposely sampled, as they served as the first point of contact for patients within the settlement. 

### Data Collection 
Data were collected through a structured quantitative assessment form administered at each health facility over one month to evaluate capacity for NTD management. The form captured facility characteristics (type, location, and bed capacity), laboratory capacity (availability of a dedicated lab, technician, and microscope), and diagnostic resources (tests for intestinal schistosomiasis and STH, alongside clinical guidelines and protocols). Reporting infrastructure was assessed through the use of the Health Management Information System and the availability of reporting forms. The survey also documented essential medicines (praziquantel, albendazole, mebendazole, and ivermectin) and laboratory supplies required for diagnosis, including slides, centrifuges, filters, syringes, staining reagents, Kato-Katz templates, and related consumables. 

### Data Analysis 
Completed questionnaires were first logged into Microsoft Excel 2016 for data entry, cleaning, and coding to ensure accuracy and consistency, after which the dataset was exported to STATA 17.0 for analysis. Facility characteristics such as facility type and distance from the town council were analyzed as categorical variables and summarized using frequencies and percentages, while the number of hospital beds was treated as a continuous variable and summarized using means and ranges. Laboratory capacity (presence of a laboratory, technician, and microscope), diagnostic resources (availability of tests and diagnostic guidelines), reporting infrastructure (HMIS use and reporting forms), essential medicines, and laboratory supplies were all analyzed as binary categorical variables (present = 1, absent = 0) and summarized descriptively using frequencies and proportions. Multi-select categories, such as diagnostic tests and laboratory consumables, were further disaggregated to present the availability of each item across the eight facilities, allowing a detailed profile of service readiness. 

### Ethics Statement 
This study was conducted in accordance with the ethical principles outlined in the Declaration of 
Helsinki. Ethics approval was obtained from the Lacor Hospital Institutional Research Ethics Committee (Reference Number: LACOR-2024-362). Written informed consent was obtained from all participants prior to data collection. Participants were informed about the study objectives, procedures, potential risks and benefits, their right to decline or withdraw at any time without any consequence, and confidentiality measures were assured. 

## Results
Table 1 summarizes the assessment of 8 health facilities within Nakivale Refugee Settlement. Of these, 6 were identified as Health Centre III (HCIII) facilities, and 2 were Health Centre IV (HCIV) facilities. Regarding geographical accessibility, 7 out of the 8 facilities were located more than 10 km from the town council, with only one Health Centre IV situated within 10 km. The average number of hospital beds varied by facility type. Health Centre III facilities reported an average of 23 hospital beds, while Health Centre IV facilities had a higher average of 50 hospital beds. 
All 8 assessed health facilities reported the presence of a dedicated laboratory and a designated lab technician for the diagnosis of Neglected Tropical Diseases. Furthermore, all 8 facilities confirmed the availability of a microscope indicating a foundational diagnostic capability

```{r, echo=FALSE}
table1 <- tribble(
  ~characteristic, ~category, ~count, ~percentage,
  "Total facilities assessed", "", "8", "N/A",

  "Facility Type", "Health Center III", "6", "75",
  "Facility Type", "Health Center IV", "2", "25",

  "Geographical Accessibility", ">10km from town council", "7", "87.5",
  "Geographical Accessibility", "Within 10km from town council", "1", "12.5",

  "Average Hospital Bed Capacity", "Health Center III", "23 beds", "N/A",
  "Average Hospital Bed Capacity", "Health Center IV", "50 beds", "N/A",

  "Laboratory Presence", "Dedicated Laboratory", "8", "100",
  "Laboratory Presence", "Designated Lab Technician", "8", "100",
  "Laboratory Presence", "Microscope availability", "8", "100"
)

table1 %>%
  kable(
    col.names = c("Characteristic", "Category", "Count", "Percentage (%)"),
    align = c("l", "l", "c", "c"),
    caption = "Table 1: Baseline Characteristics of Assessed Health Facilities (N = 8)",
    booktabs = TRUE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  add_header_above(c(" " = 2, "Summary" = 2)) %>%
  collapse_rows(columns = 1, valign = "top")
```

 
### Diagnostic Tests and Guiding Tools 
Table 2 shows characteristics regarding essential medicines for the management of STH and schistosomiasis, with the majority of facilities being well-stocked. Seven out of the 8 facilities reported having both Albendazole and Mebendazole available. One facility reported having only Albendazole. All facilities, however, lacked praziquantel and ivermectin.

The availability of specific diagnostic tests for intestinal schistosomiasis and STH showed some variation across facilities, but commonly included microscopes, dyes, and assay strips. In terms of tools to guide diagnosis, all facilities reported the presence of at least some of the listed items, such as clinical guidelines, laboratory protocols, and bench aids, though the specific combination of these tools differed between facilities. 


```{r, echo=FALSE}

table2 <- tribble(
  ~characteristic, ~item, ~facilities_with_item, ~percentage,

  "Availability of Essential Medicines", "Albendazole/Mebendazole (both)", "7", "87.5",
  "Availability of Essential Medicines", "Albendazole only", "1", "12.5",
  "Availability of Essential Medicines", "Praziquantel", "0", "0",
  "Availability of Essential Medicines", "Ivermectin", "0", "0",

  "Availability of Key Lab supplies", "Centrifuge tubes", "8", "100",
  "Availability of Key Lab supplies", "Syringes", "8", "100",
  "Availability of Key Lab supplies", "Cellophane", "0", "N/A",
  "Availability of Key Lab supplies", "Kato Katz templates", "0", "N/A",
  "Availability of Key Lab supplies", "Urine dipsticks", "7", "87.5",
  "Availability of Key Lab supplies", "Dyes", "5", "62.5",
  "Availability of Key Lab supplies", "Spatula", "3", "37.5",
  "Availability of Key Lab supplies", "Membrane filters", "2", "25",
  "Availability of Key Lab supplies", "Iodine", "4", "50",
  "Availability of Key Lab supplies", "Malachite Green", "1", "12.5",
  "Availability of Key Lab supplies", "Glycerine solution", "1", "12.5",

  "Available Diagnostic Tests for Intestinal Schistosomiasis and STH", "Clinical Guidelines", "7", "87.5",
  "Available Diagnostic Tests for Intestinal Schistosomiasis and STH", "Laboratory protocols", "6", "75",
  "Available Diagnostic Tests for Intestinal Schistosomiasis and STH", "Bench aids", "5", "62.5"
)

table2 %>%
  kable(
    col.names = c("Characteristic", "Item", "Facilities with Item", "Percentage (%)"),
    align = c("l", "l", "c", "c"),
    caption = "Table 2: Critical Gaps in Essential NTD Diagnostics and Treatment Medications",
    booktabs = TRUE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  add_header_above(c(" " = 2, "Availability" = 2)) %>%
  collapse_rows(columns = 1, valign = "top")

```

### Reporting Systems 
The assessment of reporting infrastructure revealed a notable gap. While 5 of the 8 facilities had reporting forms available for the Health Management Information System, 3 facilities reported that these forms were not available. This indicates an inconsistency in the capacity for systematic data collection and reporting on NTD diagnosis and treatment (Table 3). 

 
```{r, echo=FALSE}

table3 <- tribble(
  ~characteristic, ~category, ~facilities, ~percentage,

  "Availability of Reporting Forms", "Reporting forms for HMIS are available", "5", "62.5",
  "Availability of Reporting Forms", "Reporting forms for HMIS not available", "3", "37.5"
)

table3 %>%
  kable(
    col.names = c("Characteristic", "Category", "Facilities", "Percentage (%)"),
    align = c("l", "l", "c", "c"),
    caption = "Table 3: Inconsistent NTD Reporting Infrastructure and Geographical Accessibility Challenges",
    booktabs = TRUE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  add_header_above(c(" " = 2, "Summary" = 2)) %>%
  collapse_rows(columns = 1, valign = "top")

```

## Discussion 
This study identified four major gaps in the readiness of health facilities in Nakivale Refugee Settlement to provide comprehensive NTD services, which included inadequate diagnostic supplies despite the presence of basic laboratory infrastructure, absence of essential medicines such as praziquantel and ivermectin, weak reporting systems, and geographical inaccessibility of most facilities. These limitations constrain the effectiveness of NTD prevention, diagnosis, and management, despite some encouraging strengths such as universal availability of laboratories, technicians, and microscopes. 

Whereas all the facilities assessed reported having a designated laboratory technician and a functional microscope, critical diagnostic supplies were either inconsistently available or completely absent. When compared to other studies done in similar remote settings, the stronger baseline diagnostic infrastructure was an improvement, unlike similar studies in Tanzania and Burundi, where laboratory capacity was far more limited (40-41). However, although centrifuge tubes and syringes were available in all facilities, essential tools for WHO-recommended procedures such as urine sedimentation, centrifugation, and the Kato–Katz method (e.g., Kato-Katz templates, cellophane, and malachite green) were unavailable. This pattern mirrors findings from Tanzania, where facilities relied heavily on urine dipsticks and direct smear microscopy due to similar supply gaps (40). While urine dipsticks and direct smears are useful for rapid screening, their low sensitivity undermines accurate disease burden estimation, especially in high-risk refugee populations. 

There was a noted absence of praziquantel and ivermectin across all facilities, despite their inclusion in the WHO Model List of Essential Medicines for schistosomiasis and onchocerciasis control (41). In contrast, albendazole and mebendazole were widely available (87.5%), reflecting an imbalance in NTD drug procurement. This finding is particularly concerning given the high prevalence of schistosomiasis in refugee-hosting regions of Uganda (42). Similar stock-outs have been documented in other low-resource settings, where fragmented supply chains and donor-driven drug distribution lead to inconsistent medicine availability (43). Without praziquantel and ivermectin, facilities in Nakivale Refugee Settlement remain unable to deliver integrated, guideline-based NTD care. 

There were inconsistencies with the reporting infrastructure, with only 62.5% of facilities having HMIS reporting forms. This inconsistency limits systematic data capture, surveillance, and accountability. Similar reporting gaps have been observed in other sub-Saharan African health systems, where underreporting of NTD cases undermines programmatic planning and resource allocation (44). Strengthening reporting mechanisms is particularly critical in refugee settlements, where disease dynamics differ from national averages due to overcrowding, poor sanitation, and frequent mobility. 

With regards to facility location, 87.5% of facilities were located more than 10 km from the town council, with only one HCIV within 10 km. This distance barrier adds to the structural challenges refugees face in accessing care. Previous studies have consistently highlighted that increased distance to health facilities reduces care-seeking behaviour, increases delays, and contributes to poor adherence to long-term treatment regimens (48). For NTDs that require periodic preventive chemotherapy, such barriers undermine programme reach and effectiveness. 
This study’s strengths include its ability to identify the available diagnostic infrastructure in a refugee setting; however, the use of a single refugee settlement as a study site limits generalizability to other refugee settings. Availability of supplies and equipment was based on staff reports rather than direct observation or inventory checks, introducing a risk of reporting bias. The study also did not assess the quality of diagnostic practices or adherence to guidelines in daily use. 
Future qualitative studies could explore the reasons behind the lack of specific laboratory consumables and reporting forms, as well as the challenges faced by health facility staff in maintaining these resources. Investigating the impact of facility distance on patient access to NTD services would also provide valuable insights. 

In conclusion, this assessment shows that while Nakivale Refugee Settlement’s health facilities have a solid foundation for NTD management with labs, microscopes, and essential medicines in place, critical gaps remain. The lack of specific diagnostic supplies like Kato-Katz templates and cellophane, gaps in reporting systems, and the remote location of most facilities all pose significant challenges to effective NTD control. To address these, a regular supply of key laboratory consumables should be ensured through quarterly procurement reviews and better stock management. Strengthening and digitizing NTD reporting systems is also essential to improve surveillance and planning. Practical solutions to overcome remoteness, such as mobile outreach clinics or community-based diagnosis, should be prioritized.

Finally, future research should include qualitative studies to explore how distance and other barriers affect service use. Such insights will help tailor interventions to strengthen service delivery and accelerate progress toward NTD elimination in this vulnerable setting. 

## List of Abbreviations 
NTDs: Neglected Tropical Diseases 

PHC: Primary Healthcare 

WHO: World Health Organization 

HCIII: Health Centre III 

HCIV: Health Centre IV 

STH: Soil-Transmitted Helminths 

SDG: Sustainable Development Goal 

USAID: United States Agency for International Development 

## Declarations 
### Ethics approval and consent to participate 
Following the Declaration of Helsinki, Ethics approval for this study was obtained from the Lacor Hospital Institutional Research Ethics Committee, Reference Number [LACOR-2024-362]. Informed consent was obtained from all participants before data collection. 

### Clinical trial number: 
Not Applicable 

### Consent for publication: 
Not Applicable 

### Availability of data and materials: 
The datasets generated and/or analyzed during the current study are not publicly available due to the sensitive nature of participant data and to maintain participant anonymity, but are available from the corresponding author on reasonable request. 

### Competing interests: 
The authors declare that they have no competing interests. 

### Funding: 
This study was funded by an early-career grant from the Royal Society of Tropical Medicine and Hygiene (RSTMH). 

### Authors' contributions: 
TE conceived the study, participated in its design and coordination, and helped to draft the manuscript. AP and MPD participated in the study's design, performed the statistical analysis, and drafted the manuscript. NMS, AC, MC, SK, AP, and MPD participated in data collection and interpretation. All authors read and approved the final manuscript 

### Acknowledgements: 
The authors would like to thank the health workers in Nakivale Refugee Settlement for their participation in this study. We also extend our gratitude to Medical Teams International for their assistance during data collection. We also extend gratitude to the RSTMH Early Career Grant funding that facilitated the research process 

## References 
1.	Emeto, D. C., et al. (2021). Recognition and reporting of neglected tropical diseases by primary health care workers in Ibadan, Nigeria. Pan African Medical Journal, 38. https://doi.org/10.11604/pamj.2021.38.224.20576

2.	Molyneux, D. H., Savioli, L., & Engels, D. (2017). Neglected tropical diseases: Progress toward addressing the chronic pandemic. The Lancet, 389(10066), 312–325. https://doi.org/10.1016/S0140-6736(16)30171-4

3.	Ochola, E. A., Elliott, S. J., & Karanja, D. M. S. (2021). The impact of neglected tropical diseases (NTDs) on women’s health and wellbeing in sub-Saharan Africa: A case study of Kenya. International Journal of Environmental Research and Public Health, 18, 2180. https://doi.org/10.3390/ijerph18042180

4.	Ojja, S., et al. (2018). Prevalence, intensity and factors associated with soil-transmitted helminth infections among preschool-age children in Hoima district, rural western Uganda. BMC Infectious Diseases, 18(1), 1–12. https://doi.org/10.1186/S12879-018-3289-0

5.	World Health Organization. (2017). Integrating neglected tropical diseases into global health and development: Fourth WHO report on neglected tropical diseases. WHO.

6.	World Health Organization. (2013). Implementation tools: Package of essential noncommunicable disease interventions. WHO.

7.	World Health Organization. (2015). Investing to overcome the global impact of neglected tropical diseases. WHO.

8.	World Health Organization. (n.d.). Neglected tropical diseases. Retrieved July 1, 2025.

9.	Hotez, P. J. (2008). Neglected tropical diseases in the Anthropocene: The case for a new paradigm. PLoS Neglected Tropical Diseases, 2(10), e320.

10.	Hotez, P. J., & Molyneux, D. H. (2008). Tropical diseases in the 21st century: The neglected diseases of neglected populations. PLoS Neglected Tropical Diseases, 2(10), e320.

11.	Hotez, P. J., et al. (2007). Control of neglected tropical diseases. New England Journal of Medicine, 357(10), 1018–1027.

12.	Hotez, P. J., et al. (2006). Incorporating a rapid-impact package of neglected tropical diseases control into AIDS, tuberculosis, and malaria control efforts. PLoS Medicine, 3(5), e102.

13.	Molyneux, D. H., Hotez, P. J., & Fenwick, A. (2005). Rapid-impact interventions: Integrated control for neglected tropical diseases. PLoS Medicine, 2(11), e336.

14.	Hotez, P. J., Molyneux, D. H., & Fenwick, A. (2007). Neglected tropical diseases: The ancient afflictions of stigma and poverty. PLoS Neglected Tropical Diseases, 1(1), e11.

15.	Hotez, P. J. (2007). The global burden of neglected tropical diseases. PLoS Neglected Tropical Diseases, 1(1), e1.

16.	Kolaczinski, J. H., Onapa, A. W., Kabatereine, N. B., Ndyomugyenyi, R., Kakembo, A. S. L., & Brooker, S. (2006). Neglected tropical diseases and their control in Uganda: Situation analysis and needs assessment. Ministry of Health Uganda.

17.	Lenk, E. J., et al. (2016). Productivity loss related to neglected tropical diseases eligible for preventive chemotherapy: A systematic review. PLoS Neglected Tropical Diseases, 10(2), e0004397. https://doi.org/10.1371/journal.pntd.0004397

18.	Ortu, G., & Williams, O. (2017). Neglected tropical diseases: Long-term approaches toward sustainable disease elimination. Infectious Diseases of Poverty, 6(1), 1–12. https://doi.org/10.1186/s40249-017-0361-8

19.	Molyneux, D. H., Hotez, P. J., & Fenwick, A. (2009). Neglected tropical diseases and the Millennium Development Goals. The Lancet, 373(9670), 1570–1572.

20.	Stolk, W. A., et al. (2016). Between-country inequalities in the neglected tropical disease burden. PLoS Neglected Tropical Diseases, 10(5), e0004560. https://doi.org/10.1371/journal.pntd.0004560

21.	Hotez, P. J., et al. (2009). Neglected tropical diseases in Africa: The case for a new paradigm. PLoS Neglected Tropical Diseases, 3(9), e486.

22.	ACT to End NTDs East. (2022). Uganda. Retrieved July 4, 2025.

23.	USAID. (n.d.). Act to End NTDs | East: Uganda. Retrieved July 2, 2025.

24.	World Health Organization. (2021). Ending the neglect to attain the Sustainable Development Goals: Road map for neglected tropical diseases 2021–2030. WHO.

25.	World Health Organization. (2020). Ending the neglect to attain the Sustainable Development Goals: Road map for neglected tropical diseases 2021–2030. WHO.

26.	Hotez, P. J., & Aksoy, S. (2018). Integrating neglected tropical diseases into primary health care. The Lancet Global Health, 6(1), e2–e3.

27.	World Health Organization. (2010). Working to overcome the global impact of neglected tropical diseases. WHO.

28.	Vega, J. (2013). Universal health coverage: The post-2015 development agenda. The Lancet, 381, 179–180. https://doi.org/10.1016/S0140-6736(13)60062-8

29.	Bangert, M., et al. (2017). The cross-cutting contribution of ending neglected tropical diseases to the Sustainable Development Goals. Infectious Diseases of Poverty, 6, 73. https://doi.org/10.1186/s40249-017-0288-0

30.	World Health Organization. (2017). Ending the neglect to attain the Sustainable Development Goals. WHO.

31.	World Health Organization. (2012). Accelerating work to overcome the global impact of neglected tropical diseases. WHO.

32.	Bizimana, P., et al. (2018). Capacity gaps in health facilities for schistosomiasis and STH management in Burundi. Infectious Diseases of Poverty, 7(1), 1–9. https://doi.org/10.1186/s40249-018-0447-y

33.	Mazigo, H. D., et al. (2021). Primary health care capacity gaps in schistosomiasis management in Tanzania. BMC Health Services Research, 21(1), 1–9. https://doi.org/10.1186/s12913-021-06531-z

34.	Okello, A. L., Nsereko, V., & Ocan, M. (2016). Knowledge, attitudes, and practices of health workers on neglected tropical diseases in Uganda. Journal of Public Health in Africa, 7(1), 517.

35.	Mfinanga, S. G., Mmbaga, E. J., & Mchome, S. (2017). Knowledge, attitudes, and practices of health care workers on neglected tropical diseases in Tanzania. PLoS Neglected Tropical Diseases, 11(11), e0006009.

36.	Ministry of Health Uganda. (2020). Sustainability plan for neglected tropical diseases control program.

37.	UNHCR. (2021). Uganda refugee response.

38.	UNHCR. (n.d.). Uganda refugee statistics. Retrieved July 3, 2025.

39.	World Health Organization. (n.d.). Neglected tropical diseases. Retrieved July 1, 2025.

40.	Kihumuro, R. B., et al. (2024). Healthcare professionals’ perspectives on neglected tropical diseases in Eastern Uganda. Transactions of the Royal Society of Tropical Medicine and Hygiene, 118(12), 781–789. https://doi.org/10.1093/trstmh/trae043

41.	Krogsbøll, L. T., Jørgensen, K. J., & Gøtzsche, P. C. (2015). Screening with urinary dipsticks for reducing morbidity and mortality. Cochrane Database of Systematic Reviews, CD010007. https://doi.org/10.1002/14651858.CD010007.pub2

42.	Kioko, U., et al. (2025). Strengthening the sustainability of neglected tropical disease programs in Rwanda. PLoS Neglected Tropical Diseases, 19(8), e0012371.

43.	Olaniran, A., et al. (2022). Stock-outs of essential medicines among community health workers in LMICs. Human Resources for Health, 20(1), 58. https://doi.org/10.1186/s12960-022-00755-8

44.	Oleribe, O. O., et al. (2019). Identifying key challenges facing healthcare systems in Africa and potential solutions. International Journal of General Medicine, 12, 395–403. https://doi.org/10.2147/IJGM.S223882

